### 2020-05-13 writeup/proof_transfer_ex.tex

We first state several helper lemmas that can be viewed as corollaries of Theorem \ref{thm_model_shift}.
\begin{lemma}\label{lem_minv}[\todo{ref?}]
	Let $X\in\real^{n\times p}$ that contains i.i.d. row vectors with mean $0$ and covariance $\Sigma$.
	In the setting when $n = c p$ we have that as $p$ goes to infinity,
	\[ \bigtr{(X^{\top}X)^{-1}\Sigma} = \frac{1}{c - 1}. \]
\end{lemma}

\begin{lemma}\label{lem_cov_shift}
	\todo{restate the setting} In the setting of Theorem \ref{thm_model_shift}, we have with high probability $1-{\rm o}(1)$,
\begin{align}\label{lem_cov_shift_eq}
\tr ((X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2) = \frac{1}{n_1+n_2}\cdot \bigtr{ (a_1 M^\top M + a_2)^{-1}} +\bigo{ n^{-1/2+\epsilon}},
\end{align}
	for any constant $\epsilon>0$.
\end{lemma}
We will give the proof of Lemma \ref{lem_cov_shift} in Section \ref{sec_maintools}.

\subsection{Proof of Proposition \ref{prop_transition}}

%\paragraph{A tighter bound for a special case.} 
Under the setting of Proposition \ref{prop_transition}, we can get a bound tighter than Theorem \ref{thm_model_shift} as follows.

@@

Using Lemma \ref{lem_minv} and \ref{lem_cov_shift}, we can track the change of variance from $\hat{\beta}_t^{\MTL}$ to $\hat{\beta}_t^{\STL}$ as follows
The proof will consist of two main steps.
\begin{itemize}
	\item First, we show that $\hat{v}$ is close to $1$.
	\item Second, we plug $\hat{v}$ back into $\te(\hat{\beta}_t^{\MTL})$ to show the result.
\end{itemize}

@@

We denote
\begin{align}
	\val(w) &= n_1\left[d^2 +\left( w-1\right)^2\kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_2^{\top}X_2)^2\right] \nonumber \\
	& + n_2\left[d^2 +\left( w-1\right)^2\kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} w^2(X_1^{\top}X_1)^2\right] \nonumber\\
			& + (n_1+n_2 w^2)\sigma^2 \cdot \bigtr{(w^2X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }. \nonumber
\end{align}
Under the setting of Proposition \ref{prop_model_shift_tight}, using concentration of random vector with i.i.d. entries, we have that
$$\val(\hat{B}; w_1, w_2)= \val(w) \left( 1+\oo(1)\right)\quad \text{with probability }1-\oo(1).$$
Thus it suffices to study the behavior of $\val (w)$. For the minimizer $\hat w$ of $\val(w)$, we have a similar result as in Proposition \ref{thm_cov_shift}.

@@

Taking derivative of $\val(w)$ with respect to $w$, we obtain that 
\begin{align*}
\val'(w) \ge &~ n_1\left[ \frac{2(w-1)(2-w)}{w^5}\kappa^2 - \frac{4d^2}{w^5}\right] \tr\left[( X_1^{\top}X_1 +w^{-2}X_2^{\top}X_2)^{-2} (X_2^{\top}X_2)^2\right]   \\
+&~ n_2\left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right]\cdot \tr\left[( X_1^{\top}X_1 +w^{-2}X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \nonumber\\
		-&~ 2n_1 \frac{\sigma^2}{w^3} \cdot \bigtr{(X_1^{\top}X_1 + w^{-2}X_2^{\top}X_2)^{-2} X_1^\top X_1  } = n_1 \bigtr{(X_1^{\top}X_1 + w^{-2}X_2^{\top}X_2)^{-2} \cal A },
\end{align*}
where the matrix $\cal A$ is
\begin{align*}
\cal A :&= \left[ \frac{2(w-1)(2-w)}{w^5}\kappa^2 - \frac{4d^2}{w^5}\right](X_2^{\top}X_2)^2 + \frac{n_2}{n_1} \left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right](X_1^{\top}X_1)^2. - 2\frac{\sigma^2}{w^3}X_1^\top X_1
\end{align*}
Using the estimate \eqref{eq_isometric}, we get that $\cal A$ is lower bounded as
\begin{align*}
\cal A \succeq - \frac{4d^2}{w^5}(\sqrt{n_2}+\sqrt{p})^4 + \frac{n_2}{n_1} \left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right](\sqrt{n_1}-\sqrt{p})^4 - 2\frac{\sigma^2}{w^3}(\sqrt{n_1}+\sqrt{p})^2\succ 0,
\end{align*}
as long as 
$$w> w_1:=1 +\frac{d^2}{\kappa^2}+ \frac{\sigma^2}{\kappa^2}\frac{n_1(\sqrt{n_1}+\sqrt{p})^2}{n_2(\sqrt{n_1}-\sqrt{p})^4}+ \frac{2d^2}{\kappa^2}\frac{n_1(\sqrt{n_2}+\sqrt{p})^4}{n_2(\sqrt{n_1}-\sqrt{p})^4}.$$
Hence $\val'(w)>0$ on $(1+w_1,\infty)$, i.e. $\val(w)$ is strictly increasing for $w>1+w_1$. Hence we must have $\hat w\le 1+w_1$. Note that under our assumptions, we have $w_1=1+\OO(p^{-1})$.

Then we consider the case $w\le 1$. Taking derivative of $\val(w)$ with respect to $w$, we obtain that 
\begin{align}
	\val'(w) &\le n_1 \left[2\left( w-1\right) \kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_2^{\top}X_2)^2\right] \nonumber \\
	& + n_2\left[2d^2w +2w\left( w-1\right)(2w-1)\kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \nonumber\\
			& + 2n_2 w\sigma^2 \cdot \bigtr{(w^2X_1^{\top}X_1  + X_2^{\top}X_2)^{-2} X_2^\top X_2} = n_1 \bigtr{(w^2X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} \cal B} , \nonumber
\end{align}
where the matrix $\cal B$ is 
$$\cal B= 2\left( w-1\right) \kappa^2  (X_2^{\top}X_2)^2+\frac{n_2}{n_1}\left[2d^2w +2w\left( w-1\right)(2w-1)\kappa^2\right](X_1^{\top}X_1)^2 + 2\frac{n_2}{n_1} w\sigma^2X_2^\top X_2.$$
\cob Using the estimate \eqref{eq_isometric}, we get that $\cal A$ is lower bounded as
\begin{align*}
\cal A \succeq - \frac{4d^2}{w^5}(\sqrt{n_2}+\sqrt{p})^4 + \frac{n_2}{n_1} \left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right](\sqrt{n_1}-\sqrt{p})^4 - 2\frac{\sigma^2}{w^3}(\sqrt{n_1}+\sqrt{p})^2\succ 0,
\end{align*}
as long as 
$$w> w_1:=1 +\frac{d^2}{\kappa^2}+ \frac{\sigma^2}{\kappa^2}\frac{n_1(\sqrt{n_1}+\sqrt{p})^2}{n_2(\sqrt{n_1}-\sqrt{p})^4}+ \frac{2d^2}{\kappa^2}\frac{n_1(\sqrt{n_2}+\sqrt{p})^4}{n_2(\sqrt{n_1}-\sqrt{p})^4}.$$
Hence $\val'(w)>0$ on $(1+w_1,\infty)$, i.e. $\val(w)$ is strictly increasing for $w>1+w_1$. Hence we must have $\hat w\le 1+w_1$. Note that under our assumptions, we have $w_1=1+\OO(p^{-1})$. 
\end{proof}

>>>>

We state several helper lemmas to get a bound on the variance of $\hat{\beta}_t^{\STL}$ and $\hat{\beta}_t^{\MTL}$.
The first lemma, which is a folklore result in random matrix theory, helps determine the asymptotic limit of $\te(\hat{\beta}_t^{\STL})$, as $p$ goes to infinity.
\begin{lemma}\label{lem_minv}[\todo{ref?}]
	Let $X\in\real^{n\times p}$ that contains i.i.d. row vectors with mean $0$ and covariance $\Sigma$.
	In the setting when $n = c p$ we have that as $p$ goes to infinity,
	\[ \bigtr{(X^{\top}X)^{-1}\Sigma} = \frac{1}{c - 1}. \]
\end{lemma}

The second lemma is a corollary of Theorem \ref{thm_model_shift}.
\begin{lemma}\label{lem_cov_shift}
	Let $X_i\in\real^{n_i\times p}$ be a random matrix that contains i.i.d. row vectors with mean $0$ and variance $\Sigma_i$, for $i = 1, 2$.
	Denote by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and let $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M^{\top}M$ in decreasing order.
	When $n_1 = c_1 p$ and $n_2 = c_2 p$, we have that with high probability over the randomness of $X_1$ and $X_2$, the following equation holds
	\begin{align}%\label{lem_cov_shift_eq}
		\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2} = \frac{1}{c_1+c_2}\cdot \bigtr{ \frac{1}{p}\cdot(a_1 M^\top M + a_2)^{-1}} +\bigo{ n^{-1/2+\epsilon}},
	\end{align}
	for any constant $\epsilon>0$, where $a_1, a_2$ are solutions to the following deterministic equations:
	\begin{align}
		a_1 + a_2 = 1- \frac{1}{c_1 + c_2},~ a_1 + \sum_{i=1}^p \frac{a_1}{(c_1 + c_2)(a_1 + a_2/ \lambda_i^2)} = \frac{c_1}{c_1 + c_2}. \label{eq_a12}
	\end{align}
\end{lemma}
We will give the proof of Lemma \ref{lem_cov_shift} in Section \ref{sec_maintools}.

\subsection{Proof of Proposition \ref{prop_dist_transition}}

%\paragraph{A tighter bound for a special case.} 
Under the setting of Proposition \ref{prop_dist_transition}, we can get a bound tighter than Theorem \ref{thm_model_shift} as follows.

@@

Using Lemma \ref{lem_minv} and \ref{lem_cov_shift}, we can track the change of variance from $\hat{\beta}_t^{\MTL}$ to $\hat{\beta}_t^{\STL}$ as follows

The proof will consist of two main steps.
\begin{itemize}
	\item First, we show that $\hat{v}$ is close to $1$.
	\item Second, we plug $\hat{v}$ back into $\te(\hat{\beta}_t^{\MTL})$ to show the result.
\end{itemize}

\begin{lemma}
	In the setting of Proposition \ref{prop_dist_transition}, we have that $\abs{\hat{v} = \hat{w_1} / \hat{w_2} - 1} \le o(1)$.
\end{lemma}

\begin{proof}[Proof of Proposition \ref{prop_dist_transition}, upper bound]
\end{proof}

\begin{proof}[Proof of Proposition \ref{prop_dist_transition}, lower bound]
\end{proof}

@@

We denote
\begin{align}
	\val(w) &= n_1\left[d^2 +\left( w-1\right)^2\kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_2^{\top}X_2)^2\right] \nonumber \\
	& + n_2\left[d^2 +\left( w-1\right)^2\kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} w^2(X_1^{\top}X_1)^2\right] \nonumber\\
			& + (n_1+n_2 w^2)\sigma^2 \cdot \bigtr{(w^2X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }. \nonumber
\end{align}
Under the setting of Proposition \ref{prop_model_shift_tight}, using concentration of random vector with i.i.d. Gaussian entries, Lemma \ref{largedeviation}, we have that
$$\val(\hat{B}; w_1, w_2)= \val(w) \left( 1+\OO(p^{-1/2})\right)\quad \text{with probability }1-\oo(1).$$
Thus it suffices to study the behavior of $\val (w)$. For the minimizer $\hat w$ of $\val(w)$, we have a similar result as in Proposition \ref{thm_cov_shift}.

@@

Taking derivative of $\val(w)$ with respect to $w$, we obtain that 
\begin{align*}
\val'(w) \ge &~ n_1\left[ \frac{2(w-1)(2-w)}{w^5}\kappa^2 - \frac{4d^2}{w^5}\right] \tr\left[( X_1^{\top}X_1 +w^{-2}X_2^{\top}X_2)^{-2} (X_2^{\top}X_2)^2\right]   \\
+&~ n_2\left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right]\cdot \tr\left[( X_1^{\top}X_1 +w^{-2}X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \nonumber\\
		-&~ 2n_1 \frac{\sigma^2}{w^3} \cdot \bigtr{(X_1^{\top}X_1 + w^{-2}X_2^{\top}X_2)^{-2} X_1^\top X_1  } = n_1 \bigtr{(X_1^{\top}X_1 + w^{-2}X_2^{\top}X_2)^{-2} \cal A },
\end{align*}
where the matrix $\cal A$ is
\begin{align*}
\cal A :&= \left[ \frac{2(w-1)(2-w)}{w^5}\kappa^2 - \frac{4d^2}{w^5}\right](X_2^{\top}X_2)^2 + \frac{n_2}{n_1} \left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right](X_1^{\top}X_1)^2. - 2\frac{\sigma^2}{w^3}X_1^\top X_1
\end{align*}
Using the estimate \eqref{eq_isometric}, we get that $\cal A$ is lower bounded as
\begin{align*}
\cal A \succeq - \frac{4d^2}{w^5}(\sqrt{n_2}+\sqrt{p})^4 + \frac{n_2}{n_1} \left[\frac{2\left( w-1\right)}{w^3}\kappa^2 - \frac{2d^2}{w^3} \right](\sqrt{n_1}-\sqrt{p})^4 - 2\frac{\sigma^2}{w^3}(\sqrt{n_1}+\sqrt{p})^2\succ 0,
\end{align*}
as long as 
$$w> w_1:=1 +\frac{d^2}{\kappa^2}+ \frac{\sigma^2}{\kappa^2}\frac{n_1(\sqrt{n_1}+\sqrt{p})^2}{n_2(\sqrt{n_1}-\sqrt{p})^4}+ \frac{2d^2}{\kappa^2}\frac{n_1(\sqrt{n_2}+\sqrt{p})^4}{n_2(\sqrt{n_1}-\sqrt{p})^4}.$$
Hence $\val'(w)>0$ on $(w_1,\infty)$, i.e. $\val(w)$ is strictly increasing for $w>w_1$. Hence we must have $\hat w\le w_1$. Note that under our assumptions, we have $w_1=1+\OO(p^{-1})$.

Then we consider the case $w\le 1$. Taking derivative of $\val(w)$ with respect to $w$, we obtain that 
\begin{align}
	\val'(w) &\le n_1 \left[2\left( w-1\right) \kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_2^{\top}X_2)^2\right] \nonumber \\
	& + n_2\left[2d^2w +2w\left( w-1\right)(2w-1)\kappa^2\right]\cdot \tr\left[( w^2X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \nonumber\\
			& + 2n_2 w\sigma^2 \cdot \bigtr{(w^2X_1^{\top}X_1  + X_2^{\top}X_2)^{-2} X_2^\top X_2} = n_1 \bigtr{(w^2X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} \cal B} , \nonumber
\end{align}
where the matrix $\cal B$ is 
$$\cal B= 2\left( w-1\right) \kappa^2  (X_2^{\top}X_2)^2+\frac{n_2}{n_1}\left[2d^2w +2w\left( w-1\right)(2w-1)\kappa^2\right](X_1^{\top}X_1)^2 + 2\frac{n_2}{n_1} w\sigma^2X_2^\top X_2.$$
Using the estimate \eqref{eq_isometric}, we get that $\cal B$ is upper bounded as
\begin{align*}
\cal B \preceq - 2(1-w)\kappa^2(\sqrt{n_2}-\sqrt{p})^4 +2d^2 w\frac{n_2}{n_1}(\sqrt{n_1}+\sqrt{p})^4 +2\sigma^2 w \frac{n_2}{n_1}(\sqrt{n_2}+\sqrt{p})^2\prec 0,
\end{align*}
as long as 
$$w< w_2:=1 - \left[ \frac{\sigma^2}{\kappa^2}\frac{n_2(\sqrt{n_2}+\sqrt{p})^2}{n_1(\sqrt{n_2}-\sqrt{p})^4}+ \frac{d^2}{\kappa^2}\frac{n_2(\sqrt{n_1}+\sqrt{p})^4}{n_1(\sqrt{n_2}-\sqrt{p})^4}\right].$$
Hence $\val'(w)<0$ on $[0,w_2)$, i.e. $\val(w)$ is strictly increasing for $w<w_2$. Hence we must have $\hat w\le w_2$. Note that under our assumptions, we have $w_2=1-\OO(p^{-1})$. 
\end{proof}

### 2020-05-13 writeup/transfer.tex

Specifically, the transition threshold is derived in the following proposition.
\begin{proposition}\label{prop_transition}
	In the example described above, whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between the model distance and the noise level:
	\begin{itemize}
		\item If ${p\cdot d^2}/{\sigma^2} < \frac{2(2c-1)^2}{(c-1)(2c+1)} - o(1)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If ${p\cdot d^2}/{\sigma^2} \ge \frac{2(2c-1)^2}{(c-1)(2c+1)} + o(1)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

We obtain the following insight from Proposition \ref{prop_transition}.
First, adding the source task has the effect of reducing the variance of the estimator, independent of the model shift.
On the other hand, model shift introduces an additional bias term, which scales with $d^2$, the distance of the two task models.
Hence, the type of transfer is determined by the trade-off between the bias caused by model shift and the reduction of variance.

>>>>

Specifically, the transition threshold is derived in the following proposition.
\begin{proposition}\label{prop_dist_transition}
	In the example described above, whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between the model distance and the noise level:
	\begin{itemize}
		\item If ${p\cdot d^2}/{\sigma^2} < \frac{2(2c-1)^2}{(c-1)(2c+1)} - o(1)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If ${p\cdot d^2}/{\sigma^2} \ge \frac{2(2c-1)^2}{(c-1)(2c+1)} + o(1)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

We obtain the following insight from Proposition \ref{prop_dist_transition}.
First, adding the source task has the effect of reducing the variance of the estimator, independent of the model shift.
On the other hand, model shift introduces an additional bias term, which scales with $d^2$, the distance of the two task models.
Hence, the type of transfer is determined by the trade-off between the bias caused by model shift and the reduction of variance.
